Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-08 doc/invt.tex
small corrections

\begin{abstract}
  Wang-Landau (WL) and metadynamics algorithms
  are widely-used techniques in free energy simulations.
  %
  These techniques allow one to sample a flat distribution
  along a quantity of interest
  by constructing a bias potential on the fly,
  and to compute the negative potential of mean force (PMF),
  which is the negative of the bias potential.
  %
  However, the asymptotic error of the resulting PMF
  is determined by the magnitude of updating the bias potential,
  and in what fashion the magnitude is reduced
  over the simulation time.
  %
  In the Wang-Landau case,
  an effective prescription of determining the magnitude
  is to set it to be the inverse of the number of simulation steps.
  %
  Here, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case,
  in which the bias potential at several neighboring bins
  are updating according to a prescribed function,
  upon each visit.
  %
  Our analysis also suggests that in terms of the convergence
  the WL algorithm represents one of the best of the class
  updating schemes among more sophisticated candidates.
\end{abstract}

@@

The error function can be obtained
from Eq. \eqref{eq:T_orthonormal_cols}
as
\begin{align}
  \mathcal E
  &\equiv
  \sum_{i = 1}^n p_i \, x_i^2
  =
  \sum_{i, j, k=1}^n T_{ki} \, T_{kj} \, x_i \, x_j
  =
  \sum_{k = 1}^n y_k^2.
  \label{eq:y2_sum}
\end{align}
%
If we arrange $y_k$ such that
the mode $y_1 = \sum_{j=1}^n p_j \, x_j$,
(corresponding to $\lambda_1 = 1$),
then $y_1$ vanishes identically
by Eq. \eqref{eq:px_sum},
and the sum of $y_k$ in Eq. \eqref{eq:y2_sum}
can start from $k = 2$.

@@

We can find an analytical expression
of the optimal value of $\lambda$
in the limit of large $n$.
%
In this case,
we convert Eq.
\eqref{eq:y2_sum_perfect_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$$
\begin{aligned}
\mathcal E
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, m_1 \, \cos^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, m_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, m_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{The integral is evaluated by contour integration.
  %
$$
\begin{aligned}
\mathcal E
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, m_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - m_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, m_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is the unit circle
in the complex plane of $z$.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, m_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, m_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, m_1 - \lambda}}
{\sqrt{8 m_1}}.
$$
Thus,
$$
\begin{aligned}
  \mathcal E
&=
\frac{n}{\lambda \, t}
\left(
 \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - m_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 m_1 - \lambda - 4 m_1 z^2) }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ \left(1 - m_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
 { 2 - 4 m_1 - \lambda - 4 m_1 z_{\pm}^2 }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 m_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, m_1} { 1 - 2 \, m_1 },
\label{eq:lambda_tridiag}
\end{equation}
%
and the error in the optimal is given by
%
\begin{equation}
\mathcal E
=
\frac{n}{t}
\left(
  1+ \frac{2 \, m_1^2}{1-4 \, m_1}
\right).
\label{eq:error_tridiag}
\end{equation}
%
From Eq. \eqref{eq:error_tridiag},
it is clear that the error
increases rapidly with magnitude of $m_1$,
no matter its sign,
and optimal value is given by $m_1 = 0$,
which is the single-bin update scheme.

@@

For the Gaussian wave,
$m(x) \propto \exp(-\frac{1}{2} x^2/\sigma^2)$.
%
If $\Delta x \ll \sigma \ll 1$,
we can extended
the upper limit of the integrals
in Eqs. \eqref{eq:lambda_int}
and \eqref{eq:mx_normalization}
to infinity, and
%
\begin{equation}
m(x)
=
\frac{            1            }
     { \sqrt{ 2 \pi \sigma^2 } }
%
\exp\left(
      -\frac{       x^2     }
            { 2 \, \sigma^2 }
    \right),
\end{equation}
%
%
The eigenvalues are given by
%
\begin{equation}
\lambda_k
=
\exp\left(
      -\frac{ (k - 1)^2 \, \sigma^2 }
            {           2           }
    \right).
\label{eq:lambda_Gaussian}
\end{equation}
%
The eigenvalues are positive,
which means that the updating scheme
is stable.

However,
as $\sigma/\Delta x \gg 1$,
the smallest eigenvalue
%
$$
\lambda_n
\approx
\exp\left(
      -\frac{ n^2 \, \sigma^2 }
            {        2        }
    \right)
=
\exp\left[
      -\frac{ \pi^2 }{ 2 }
      \left(
        \frac{  \sigma }
             { \Delta x }
      \right)^2
    \right]
.
$$
%
is exponentially small.
%
This means that it will be difficult
to use the optimal value of $\lambda$
given by Eq. \eqref{eq:optimal_lambda_approx},
which leads to a value
$\lambda < 2 \, \lambda_n$
that can be too small in practice.
%
In other words,
with a reasonable $\lambda$,
the error of the last few fluctuation modes
will always decay suboptimally,
accordingly to Eq. \eqref{eq:y2_sum_perfect_invt}.
%
However, since these modes
represent high-frequency fluctuations,
they may be unimportant.
%
Thus,
we may truncate the sum of the error function
in Eq. \eqref{eq:y2_sum_perfect_invt}
at some $k_{\max}$,
which reflects a minimal length scale
$l_{\min} = \Delta x \, (n /k_{\max}) = \pi/k_{\max}$,
that is greater than or
at least comparable to $\sigma$.
%
Then, the optimal $\lambda$ is given by
$$
\lambda
\approx
\lambda_{ k_{\max} }
\approx
\exp\left(
      -\frac{ k_{\max}^2 \, \sigma^2 }
            {           2            }
    \right)
=
\exp\left(
      -\frac{  \pi^2 \, \sigma^2    }
            {    2 \, l_{\min}^2  }
    \right).
$$

>>>>

\begin{abstract}
  Wang-Landau (WL) and metadynamics algorithms
  are widely-used techniques in free energy simulations.
  %
  These techniques allow one to sample a flat distribution
  along a quantity of interest
  by adaptively constructing a bias potential,
  as the negative potential of mean force (PMF).
  %
  However, the asymptotic error of the resulting PMF
  is determined by the magnitude of updating the bias potential,
  and in what fashion the magnitude is reduced
  over the simulation time.
  %
  In the Wang-Landau case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  Here, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case,
  in which the bias potential at several neighboring bins
  are updated according to a predetermined window function,
  upon each visit.
  %
  Our analysis also suggests that in terms of the convergence
  the WL algorithm represents one of the best of the class
  updating schemes among more sophisticated candidates.
\end{abstract}

@@

The error function can be obtained
from Eq. \eqref{eq:T_orthonormal_cols}
as
\begin{align}
  \mathcal E
  &\equiv
  \sum_{i = 1}^n p_i \, x_i^2
  =
  \sum_{i, j, k=1}^n T_{ki} \, T_{kj} \, x_i \, x_j
  =
  \sum_{k = 1}^n y_k^2.
  \label{eq:y2_sum}
\end{align}
%
We note that there is one mode, say
$y_1 = \sum_{j=1}^n p_j \, x_j$
(corresponding to $\lambda_1 = 1$),
vanishes identically by Eq. \eqref{eq:px_sum}.
%
Thus, we can start the sum in Eq. \eqref{eq:y2_sum}
from $k = 2$.

@@

We can find an analytical expression
of the optimal value of $\lambda$
in the limit of large $n$.
%
In this case,
we convert Eq.
\eqref{eq:y2_sum_perfect_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$$
\begin{aligned}
\mathcal E
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, m_1 \, \cos^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, m_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, m_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{The integral is evaluated by contour integration.
  %
$$
\begin{aligned}
\mathcal E
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, m_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - m_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, m_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is the unit circle
in the complex plane of $z$.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, m_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, m_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, m_1 - \lambda}}
{\sqrt{8 m_1}}.
$$
Thus,
$$
\begin{aligned}
  \mathcal E
&=
\frac{n}{\lambda \, t}
\left(
 \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - m_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 m_1 - \lambda - 4 m_1 z^2) }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ \left(1 - m_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
 { 2 - 4 m_1 - \lambda - 4 m_1 z_{\pm}^2 }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 m_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, m_1} { 1 - 2 \, m_1 },
\label{eq:lambda_tridiag}
\end{equation}
%
and the error in the optimal is given by
%
\begin{equation}
\mathcal E
=
\frac{n}{t}
\left(
  1+ \frac{2 \, m_1^2}{1-4 \, m_1}
\right).
\label{eq:error_tridiag}
\end{equation}
%
%From Eq. \eqref{eq:error_tridiag}, it is clear that
Clearly, the error
increases rapidly with magnitude of $m_1$,
no matter its sign,
and optimal value is given by $m_1 = 0$,
which is the single-bin update scheme.

@@

For the Gaussian wave,
$m(x) \propto \exp(-\frac{1}{2} x^2/\sigma^2)$.
%
If $\Delta x \ll \sigma \ll 1$,
we can extended
the upper limit of the integrals
in Eqs. \eqref{eq:lambda_int}
and \eqref{eq:mx_normalization}
to infinity, and
%
\begin{equation}
m(x)
=
\frac{            1            }
     { \sqrt{ 2 \pi \sigma^2 } }
%
\exp\left(
      -\frac{       x^2     }
            { 2 \, \sigma^2 }
    \right),
\end{equation}
%
%
The eigenvalues are given by
%
\begin{equation}
\lambda_k
=
\exp\left(
      -\frac{ (k - 1)^2 \, \sigma^2 }
            {           2           }
    \right).
\label{eq:lambda_Gaussian}
\end{equation}
%
The eigenvalues are all positive,
suggesting stability of the updating scheme.

However,
as $\sigma/\Delta x \gg 1$,
the smallest eigenvalue
%
$$
\lambda_n
\approx
\exp\left(
      -\frac{ n^2 \, \sigma^2 }
            {        2        }
    \right)
=
\exp\left[
      -\frac{ \pi^2 }{ 2 }
      \left(
        \frac{  \sigma }
             { \Delta x }
      \right)^2
    \right]
.
$$
%
is exponentially small.
%
This means that
the optimal value of $\lambda$
given by Eq. \eqref{eq:optimal_lambda_approx}
would be too small in practice,
for it requires $\lambda < 2 \, \lambda_n$.
%
In other words,
with a reasonable $\lambda$,
the error of the last few fluctuation modes
will always decay suboptimally,
accordingly to Eq. \eqref{eq:y2_sum_perfect_invt}.
%
However, since these modes
represent high-frequency fluctuations,
they may be unimportant.
%
Thus,
we can truncate the sum of the error function
in Eq. \eqref{eq:y2_sum_perfect_invt}
at some $k_{\max}$,
which corresponds to a minimal length scale
$l_{\min} = \Delta x \, (n /k_{\max}) = \pi/k_{\max}$
comparable to $\sigma$.
%
Then, the optimal $\lambda$ is given by
$$
\lambda
\approx
\lambda_{ k_{\max} }
\approx
\exp\left(
      -\frac{ k_{\max}^2 \, \sigma^2 }
            {           2            }
    \right)
=
\exp\left(
      -\frac{  \pi^2 \, \sigma^2    }
            {    2 \, l_{\min}^2  }
    \right).
$$
